\section{Our Contribution}
\label{sec:contribution}

We define a family of dependency graphs from \module{Mathlib} and analyze them as a multi-layer network. Figure~\ref{fig:three-views-main} illustrates the three primary layers on a concrete example. When a proof cites a premise, the kernel records a directed edge between the two declarations (Figure~\ref{fig:three-views-main}b, red identifiers). The collection of all such edges forms the \emph{declaration dependency graph}~$G_{\mathrm{thm}}$ (Definition~\ref{def:thm-graph}), whose 308,129 nodes and 8.4 million edges encode the logical content of the library. Declarations reside in source files (Figure~\ref{fig:three-views-main}a); each file constitutes a module, and each \texttt{import} statement between files produces an edge in the \emph{module import graph}~$G_{\mathrm{module}}$ (Definition~\ref{def:module-graph}). In the isometric view (Figure~\ref{fig:three-views-main}c), the blue cards are modules: of six declaration-level edges, only one stays within its module, while the remaining five cross file boundaries and collapse into blue import arrows at the module level. A third organizational layer, orthogonal to the file hierarchy, is the human-chosen naming taxonomy: the \emph{namespace graphs}~$G_{\mathrm{ns}}^{(k)}$ (Definition~\ref{def:ns-graph}) aggregate declarations by their dotted-name prefix (e.g., all declarations under \ns{Mathlib.Topology}). Together with the module tree~$T$ (Definition~\ref{def:module-tree}), these layers range from fine-grained mathematical logic to coarse-grained human organization.

\begin{figure}[ht]
\centering
% ── Panel (a): File-system tree (top left) ──
\begin{subfigure}[t]{0.35\textwidth}
\centering
\vspace{0pt}
\begin{forest}
  for tree={
    font=\scriptsize\ttfamily,
    text=blue!60!black,
    grow'=0,
    child anchor=west,
    parent anchor=south,
    anchor=west,
    calign=first,
    edge path={
      \noexpand\path [draw, black, \forestoption{edge}]
        (!u.south west) +(7.5pt,0) |- (.child anchor)\forestoption{edge label};
    },
    before typesetting nodes={
      if n=1 {insert before={[,phantom]}} {}
    },
    fit=band,
    before computing xy={l=12pt},
  }
[Data/
  [Nat/
    [Basic.lean
      [\textcolor{red!70!black}{add\_comm}]
      [\textcolor{red!70!black}{add\_left\_comm}]
    ]
    [Defs.lean
      [\textcolor{red!70!black}{succ\_eq\_add\_one}]
      [\textcolor{red!70!black}{zero\_add}]
    ]
  ]
]
\end{forest}
\caption{File-system hierarchy~$T$ (\textcolor{blue!60!black}{blue}), extended to the declaration level.}
\label{fig:tree-view-main}
\end{subfigure}%
\hfill
% ── Panel (b): Source code (top right) ──
\begin{subfigure}[t]{0.60\textwidth}
\centering
\vspace{0pt}
\raggedright\scriptsize
\textsf{\bfseries\color{blue!60!black}Basic.lean}\\[2pt]
\ttfamily
\textbf{import} \textcolor{blue!60!black}{Init.Prelude}\\[3pt]
\textbf{theorem} \textcolor{red!70!black}{add\_comm} (n m : Nat)\\
\hspace{1.5em}: n + m = m + n := \textbf{by}\\
\hspace{1.5em}induction n <;> simp [\textcolor{red!70!black}{Eq.refl}]\\[3pt]
\textbf{theorem} \textcolor{red!70!black}{add\_left\_comm} (n m k : Nat)\\
\hspace{1.5em}: n + (m + k) = m + (n + k) := \textbf{by}\\
\hspace{1.5em}rw [\textcolor{red!70!black}{add\_comm}]; simp [\textcolor{red!70!black}{Eq.refl}]\\[6pt]
\textsf{\bfseries\color{blue!60!black}Defs.lean}\\[2pt]
\ttfamily
\textbf{import} \textcolor{blue!60!black}{Data.Nat.Basic}\\
\textbf{import} \textcolor{blue!60!black}{Init.Prelude}\\[3pt]
\textbf{theorem} \textcolor{red!70!black}{succ\_eq\_add\_one} (n : Nat)\\
\hspace{1.5em}: succ n = n + 1 := \textbf{by}\\
\hspace{1.5em}rw [\textcolor{red!70!black}{add\_comm}]; exact \textcolor{red!70!black}{Eq.refl} \_\\[3pt]
\textbf{theorem} \textcolor{red!70!black}{zero\_add} (n : Nat)\\
\hspace{1.5em}: 0 + n = n := \textbf{by}\\
\hspace{1.5em}induction n <;> exact \textcolor{red!70!black}{Eq.refl} \_
\caption{Proof source code: each \textcolor{red!70!black}{red identifier} in a proof term generates a dependency edge in~(c).}
\label{fig:source-view-main}
\end{subfigure}

\par\vspace{4pt}

% ── Panel (c): Isometric dependency graph (full width, bottom) ──
\begin{subfigure}[b]{\textwidth}
\centering
\begin{tikzpicture}[
  every node/.style={font=\scriptsize\ttfamily, inner sep=2pt},
  dot/.style={circle, fill=red!70!black, minimum size=3.5pt, inner sep=0pt},
  dep/.style={->, >=Stealth, red!70!black, semithick},
  depext/.style={->, >=Stealth, red!70!black, semithick, dashed},
  modimp/.style={->, >=Stealth, blue!60!black, thick},
]
  % ── Card 3 (top, back): module Defs ──
  \filldraw[fill=blue!8, draw=blue!50!black, semithick, line join=round]
    (0.2,4.7) -- (10.2,4.7) -- (11.4,5.09) -- (1.4,5.09) -- cycle;
  \filldraw[fill=blue!18, draw=blue!50!black, semithick, line join=round]
    (0.2,4.7) -- (10.2,4.7) -- (10.2,4.57) -- (0.2,4.57) -- cycle;
  \node[font=\footnotesize\sffamily, blue!60!black, anchor=west] at (10.4,4.63)
    {\textnormal{\module{Data.Nat.Defs}}};
  \node[dot] (sea) at (3.5,4.88) {};
  \node[font=\footnotesize\ttfamily, red!70!black, anchor=west] at (3.6,4.88) {succ\_eq\_add\_one};
  \node[dot] (za)  at (7.5,4.88) {};
  \node[font=\footnotesize\ttfamily, red!70!black, anchor=west] at (7.6,4.88) {zero\_add};

  % ── Card 2 (middle): module Basic ──
  \filldraw[fill=blue!8, draw=blue!50!black, semithick, line join=round]
    (0,2.5) -- (10.0,2.5) -- (11.2,2.89) -- (1.2,2.89) -- cycle;
  \filldraw[fill=blue!18, draw=blue!50!black, semithick, line join=round]
    (0,2.5) -- (10.0,2.5) -- (10.0,2.37) -- (0,2.37) -- cycle;
  \node[font=\footnotesize\sffamily, blue!60!black, anchor=west] at (10.2,2.43)
    {\textnormal{\module{Data.Nat.Basic}}};
  \node[dot] (ac)  at (3.3,2.68) {};
  \node[font=\footnotesize\ttfamily, red!70!black, anchor=east] at (3.2,2.68) {add\_comm};
  \node[dot] (alc) at (7.3,2.68) {};
  \node[font=\footnotesize\ttfamily, red!70!black, anchor=west] at (7.4,2.68) {add\_left\_comm};

  % ── Card 1 (bottom, front): module Init.Prelude ──
  \filldraw[fill=blue!8, draw=blue!50!black, semithick, line join=round]
    (1.0,0.15) -- (8.0,0.15) -- (8.9,0.46) -- (1.9,0.46) -- cycle;
  \filldraw[fill=blue!18, draw=blue!50!black, semithick, line join=round]
    (1.0,0.15) -- (8.0,0.15) -- (8.0,0.03) -- (1.0,0.03) -- cycle;
  \node[font=\footnotesize\sffamily, blue!60!black, anchor=west] at (8.2,0.09)
    {\textnormal{\module{Init.Prelude}}};
  \node[dot] (er)  at (4.8,0.3) {};
  \node[font=\footnotesize\ttfamily, red!70!black, anchor=west] at (4.9,0.3) {Eq.refl};


  % ── Module import dots & arrows (blue = $G_{\mathrm{module}}$) ──
  \node[circle, fill=blue!60!black, minimum size=5.5pt, inner sep=0pt] (md) at (9.8, 4.88) {};
  \node[circle, fill=blue!60!black, minimum size=5.5pt, inner sep=0pt] (mb) at (10.0, 2.68) {};
  \node[circle, fill=blue!60!black, minimum size=5.5pt, inner sep=0pt] (mo) at (7.2, 0.3) {};
  \draw[modimp] (md) to[out=-85,in=85] (mb);
  \draw[modimp] (mb) to[out=-70,in=40] (mo);
  \draw[modimp] (md) to[out=-50,in=55] (mo);

  % ── Dependency arrows (red = mathematical logic) ──
  \draw[dep] (alc) -- (ac);
  \draw[depext] (sea) to[out=-88,in=92] (ac);
  \draw[depext] (ac)  to[out=-88,in=105] (er);
  \draw[depext] (alc) to[out=-88,in=60] (er);
  \draw[depext] (sea) to[out=-85,in=120] (er);
  \draw[depext] (za)  to[out=-85,in=45] (er);

\end{tikzpicture}
\caption{An isometric view of $G_{\mathrm{thm}}$ layered by module. Each \textcolor{blue!60!black}{blue card} represents a module file; \textcolor{red!70!black}{red nodes} are declarations. Solid \textcolor{red!70!black}{red arrows} show intra-module edges; dashed \textcolor{red!70!black}{red arrows} cross module boundaries. \textcolor{blue!60!black}{Blue arrows} on the right show the corresponding module-level imports from $G_{\mathrm{module}}$.}
\label{fig:dep-view-main}
\end{subfigure}
\caption{Three views of the same declarations. (a)~The file-system hierarchy~$T$ (\textcolor{blue!60!black}{blue}), extended to the declaration level. (b)~Proof source code: each \textcolor{red!70!black}{red identifier} creates a dependency edge in~(c). (c)~An isometric view of~$G_{\mathrm{thm}}$ layered by module; of six dependency edges, only one stays within its module. \textcolor{blue!60!black}{Blue arrows} on the right show the corresponding module-level imports from $G_{\mathrm{module}}$. All names have the \texttt{Nat.}\ prefix removed.}
\label{fig:three-views-main}
\end{figure}
We then define a suite of graph decompositions (Appendix~\ref{sec:graph-definitions}, Table~\ref{tab:decompositions}): edge-level partitions (statement/proof, explicit/synthesized), mechanism-specific subgraphs (typeclass instances, coercions, structure inheritance, auto-derived edges), module-level projections (visibility, build, import utilization), and node-level attributes (definitional height, tactic usage, universe polymorphism). These definitions serve as an invitation to study formalized mathematics as an interconnected system whose global topology reveals organizational patterns and infrastructure dependencies that no local reading of proofs can expose.

To operationalize this perspective, we apply a suite of standard network theoretical metrics to each of the three dependency layers. For the module import graph we compute degree distributions, DAG depth profiles, centrality rankings, and community structure; for the declaration graph we add edge classification (statement vs.\ proof dependencies, explicit vs.\ synthesized) and robustness under node removal; for the namespace graphs we measure cross-namespace coupling, containment decay, and alignment with dependency-based communities. Together, these measurements establish a quantitative baseline for the macroscopic structure of formalized mathematics produced by humans, against which future changes can be tracked: as AI-assisted formalization grows, shifts in these metrics will indicate whether the organizational patterns we observe are artifacts of human authorship or intrinsic to the mathematics. This baseline also surfaces actionable engineering targets, including low import utilization (median 1.6\%) that motivates finer-grained import mechanisms, a 153-layer critical path that constrains parallel compilation, and high-PageRank hub declarations whose instability would trigger costly recompilation cascades (Appendix~\ref{sec:graph-definitions}; Appendices~\ref{app:module-detail} through \ref{app:ns-detail}).

The dependency networks we formalize predate our definitions. They constitute objective structures whose information content has already been acted upon by the Lean ecosystem, albeit without explicit network theoretical framing. What our formalization provides is not the network itself, but rather the \emph{explicit representation} that makes these correspondences visible, measurable, and comparable. Our empirical findings organize into three process traces, representing measurable imprints of human organizational practice embedded within the mathematical product, which we summarize below. In cases where the Lean ecosystem has already responded to a structural pattern, we identify the corresponding engineering tool. Distinguishing actionable targets from inherent structure is itself a contribution of network analysis.

Our data points are drawn from a cross-version comparison at four Mathlib snapshots (\texttt{b763c21}, January 2024; \texttt{13cef76}, June 2024; \texttt{a3cf05e}, June 2025; \texttt{534cf0b}, February 2026), confirming that engineering responses have measurable network-level effects even before the network is formally named. Full temporal network analysis is deferred to future work (\S\ref{sec:temporal-graphs}). 

\paragraph{Finding 1: Organizational divergence.}
The dual human-designed organizational systems of file layout and namespace naming exhibit strong internal coherence, yet they diverge fundamentally from the logical dependency structure revealed by network analysis. Concurrently, the type system of the proof assistant introduces a layer of tooling infrastructure that lacks any traditional mathematical counterpart.

\emph{Organizational alignment.}
The Normalized Mutual Information (NMI) score reaches $0.71$ between namespaces and modules, reflecting strong agreement among human-organized structures. Conversely, the score drops to $0.34$ between namespaces and dependency-based communities, highlighting a stark divergence between human taxonomy and underlying logic (\S\ref{sec:sa-community}). Cross-namespace edges constitute $50.9\%$ of dependencies at the declaration level (\S\ref{sec:thm-cross-namespace}), substantially exceeding the $37.1\%$ observed at the module level (\S\ref{sec:cross-namespace}). Structural containment also dissipates rapidly across hierarchies: namespace containment drops from $48.9\%$ to roughly $14\%$ by depth~$2$ (\S\ref{sec:containment-decay}), while module containment falls from $97.5\%$ to $8.8\%$ by depth~$4$ (\S\ref{sec:module-containment}).

\emph{Hub structure.}
The network's hub structure naturally bifurcates into two distinct layers: a \emph{language infrastructure} layer that manages coercions and equality reasoning, and a \emph{mathematical infrastructure} layer that grounds core concepts such as \decl{CategoryTheory.Category}, \decl{Real}, and \decl{TopologicalSpace} (\S\ref{sec:thm-types}). This foundational infrastructure is heavily automated; Lean's elaborator records whether each dependency edge was explicitly written by the author or silently inserted during type inference. The \emph{synthesis ratio} $\sigma = |E_{\mathrm{synth}}|/|E_{\mathrm{thm}}|$ (Definition~\ref{def:synth-partition}) reaches $74.2\%$ ($6{,}257{,}832$ synthesized vs.\ $2{,}178{,}534$ explicit edges), meaning barely a quarter of the dependency graph reflects human-authored reasoning.

\emph{Tool infrastructure decompositions.}
Lean's typeclass resolution mechanism generates the typeclass instance graph~$G_{\mathrm{tc}}$, comprising $1{,}817$ classes and $29{,}952$ instance edges (Definition~\ref{def:tc-graph}). Supporting this ecosystem are several specialized subgraphs. The coercion graph~$G_{\mathrm{coe}}$ maps implicit type conversions and exhibits a diameter of~$4$ (Definition~\ref{def:coercion-graph}), while the structure inheritance graph~$G_{\mathrm{ext}}$ captures $1{,}535$ \texttt{extends} relationships (Definition~\ref{def:extends-graph}). Together with automatically derived instances (Definition~\ref{def:deriving-edges}), these elements constitute a completely tool-generated fragment of the broader mathematical graph~$G_{\mathrm{thm}}$.

\emph{Compression.}
This layered architecture compresses $308{,}129$ declarations and $8{,}436{,}366$ edges down to $7{,}563$ modules and $23{,}484$ imports, roughly $40\times$ for nodes and $360\times$ for edges (\S\ref{sec:module-declaration}). A single \texttt{import Mathlib.Topology.Basic} abstracts over hundreds of declaration-level dependencies. Tracking this ratio over time offers a practical diagnostic: a declining compression factor signals that modules are growing too large or too interconnected for contributors to navigate effectively.

\paragraph{Finding 2: Structural redundancy.}
Lean's module system operates at file granularity: an \texttt{import} statement pulls in an entire source file regardless of how many declarations the importer actually uses. This coarse granularity produces three measurable consequences in the network. First, the module import graph contains a layer of intentional redundancy that developers maintain for ergonomic reasons (refactoring convenience, extensibility, instance availability) rather than for logical necessity. The \texttt{shake} linter~\cite{baanen2025} manages this redundancy within $G_{\mathrm{module}}$, keeping the mean imports per file stable at ${\sim}3.1$ even as the library doubled from $3{,}893$ to $7{,}563$ files. The \texttt{\#redundant\_imports} command~\cite{importgraph} separately detects $17.5\%$ transitive redundancy ($4{,}122$ edges; \S\ref{sec:transitive-reduction}) that \texttt{shake} intentionally preserves. Second, import utilization (the fraction of an imported module's declarations actually used by the importer, Definition~\ref{def:import-util}) stands at a median of $1.6\%$, quantifying the gap between file-level and declaration-level granularity. Third, the \texttt{public}/\texttt{private} import distinction~\cite{lean4modules,lean4v4200,lean4v4220,lean4v4250} reshapes the transitive visibility captured by $G_{\mathrm{mod}}^{\mathrm{pub}}$ (Definition~\ref{def:visibility-graph}), representing the community's first step toward finer-grained visibility control; at our snapshot, $98.8\%$ of imports remain \texttt{public} ($287$ \texttt{private}), indicating that this migration is in its earliest stage.

\paragraph{Finding 3: Semantic flattening.}
The three dependency layers exhibit strikingly different topologies, where human-organized structures are significantly deeper than logical ones. While formalization preserves traditional mathematical labels, it flattens their semantic meaning; consequently, the dependency structure is driven primarily by the process of proving rather than the content of the assertions. The depth profiles span $153$, $84$, and $8$ layers across the three respective graphs (\S\ref{sec:dag-depth}, \S\ref{sec:thm-dag-depth}, \S\ref{sec:ns-dag-depth}). The parallel scheduler in \texttt{lake} exploits this directed acyclic graph structure (Definition~\ref{def:build-graph}), managing a critical path that traverses $161$ modules with a parallelism ratio of $22.4\times$. Edge classification (Definition~\ref{def:stmt-proof-dep}) further emphasizes the dominance of proofs: $43.9\%$ of dependencies arise exclusively from proofs, compared to only $8.1\%$ from statements and $48.0\%$ from both. At the network's periphery, $44.8\%$ of all theorems act as terminal nodes that are never cited within \module{Mathlib} (\S\ref{sec:thm-leaves}), indicating that their consumers are inherently external. Beneath this visible topology lies definitional height (Definition~\ref{def:def-height}), which tracks the longest chain of nested \texttt{unfold} steps required to reach primitives. Peaking at $60$ steps with a median of $7$, this represents a hidden fragility metric highly suitable for continuous integration monitoring. At the implementation level, tactic usage (Definition~\ref{def:tactic-usage}) across $79{,}910$ proofs concentrates on \texttt{rw} ($16.9\%$), \texttt{simp} ($11.2\%$), and \texttt{exact} ($10.4\%$), displaying notable variations across namespaces that reflect distinct methodologies. Finally, to better isolate the human-authored dependency structure for downstream tasks like AI premise selection, we recommend using the explicit subgraph (Definition~\ref{def:synth-partition}), which filters out compiler-generated edges.