\section{Previous and related work}
\paragraph{Software dependency networks.}
Dependency graphs arise naturally in software ecosystems: package managers such as npm, PyPI, and Maven define directed acyclic graphs in which each package declares its requirements~\cite{decan2019,labelle2004}. In the formal mathematics setting, Blanchette et al.~\cite{blanchette2015} mine structural properties of the Isabelle AFP, making it the closest methodological predecessor to our work, and Bancerek et al.~\cite{bancerek2018} document the MML's role and evolution. Studies of these ecosystems have revealed heavy-tailed degree distributions, small-world connectivity, and fragility under targeted removal of hub packages. Our analysis applies a similar methodology to a formal mathematics library, a setting in which every edge carries a stronger guarantee than ``package~$A$ requires package~$B$ at build time.'' In a type-checked library, premise-level dependencies are mechanically certified: if a constant appears in a proof term, the kernel enforces that dependency. Module-level imports, by contrast, are developer-declared and can be redundant relative to reachability ($17.5\%$ are transitively implied, measured after Mathlib's \texttt{shake} linter has already pruned certain redundant imports in CI). The graph thus records not merely a build requirement, but a semantic relationship.

\paragraph{\module{Mathlib} growth and maintenance.}
Van Doorn, Ebner, and Lewis~\cite{vandoorn2020} described the engineering practices that sustain \module{Mathlib} as a collaboratively maintained library, addressing linting, documentation, and refactoring workflows. Ringer et al.~\cite{ringer2019} provide a canonical survey of proof engineering, noting that ``there is still relatively little work describing tools and best practices''; our contribution fills part of this gap. Commelin and Topaz~\cite{commelin2023} proposed abstraction boundaries and specification-driven development as organizing principles for formal mathematics. More recently, Commelin et al.~\cite{baanen2025} documented the challenges of scaling \module{Mathlib}, including build times, namespace reorganization, and technical debt. On the tooling side, Massot's \textsc{leanblueprint} system~\cite{massot2020leanblueprint} provides infrastructure for managing large formalization projects: mathematicians manually annotate dependency relations between statements in \LaTeX{} using \verb|\uses{}| tags, producing a dependency graph that tracks formalization progress. This approach has been adopted by several major projects, including the Liquid Tensor Experiment~\cite{lean_liquid} and the ongoing formalization of Fermat's Last Theorem~\cite{flt_lean}. Zhu et al.~\cite{zhu2025leanarchitect} complement this with \textsc{LeanArchitect}, which automatically infers dependency information from Lean source code, eliminating the manual duplication between informal and formal representations. Notably, their work includes AI integration: in a case study, the automatically extracted dependency graph guides an AI prover (Achim et al.'s Aristotle system~\cite{achim2025aristotle}) to formalize theorems in topological order. Our work differs from both in scope and purpose: \textsc{leanblueprint} and \textsc{LeanArchitect} extract dependency graphs to manage individual formalization projects; we analyze the global dependency structure of the entire \module{Mathlib} library as a network, applying tools from network science (degree distributions, community detection, robustness analysis) to characterize structural patterns that no single project would reveal. Avigad et al.~\cite{avigad2025anatomy} describe the practitioner experience of using proof assistants; Klein et al.~\cite{klein2012} address managing large-scale proofs in Isabelle, including tools for moving lemmas across modules. These works address \module{Mathlib}'s maintenance from a software engineering perspective. Our analysis complements theirs by providing a network-theoretic diagnostic of the same challenges they address qualitatively: we show, for example, that the $17.5\%$ import redundancy rate (Definition~\ref{def:redundancy-rate}, measured post-\texttt{shake}) and $0.107$ module cohesion quantify the development ergonomics overhead that maintenance practices must manage.

\paragraph{Machine learning for formal mathematics.}
The graph structure of \module{Mathlib} has been exploited in machine learning, particularly for premise selection. Yang et al.~\cite{leandojo} introduced the LeanDojo benchmark, extracting proof traces from \module{Mathlib} that implicitly define a dependency graph. Bemberek et al.~\cite{bemberek2025} constructed a heterogeneous dependency graph from this benchmark and applied relational graph convolutional networks for premise selection, achieving a $25\%$ improvement over text-only baselines. Lu et al.~\cite{leanfinder} built a directed acyclic graph of Lean~4 statements with dependency edges to support semantic search. On the knowledge-representation side, the EpiK Protocol~\cite{epik2024} constructed a knowledge graph from Lean~4 and \module{Mathlib}, applying embedding models for link prediction. Uskuplu and Moss~\cite{knowtex2025} argued that dependency graphs should become a standard feature of mathematical writing. On the extraction side, our analysis draws on several complementary tools: \texttt{lean4export}~\cite{lean4export} dumps every declaration in the Lean environment; \texttt{lean-training-data}~\cite{lean_training_data} extracts the premises invoked in each proof; \texttt{importGraph}~\cite{importgraph} produces the module-level import graph; and \texttt{jixia}~\cite{jixia2024} exposes per-declaration metadata such as tactic usage and definitional heights. These systems consume the dependency graph as training data or as a retrieval index. We study the graph itself, not as input to a downstream task, but as an object of independent structural interest. Our structural analysis (two-layer hub structure, community detection) could inform premise selection by quantifying how much of the search space is language infrastructure versus mathematical content.

\paragraph{Complex networks.}
The analytical tools we employ are basic and standard in network science. Heavy-tailed network theory~\cite{barabasi1999}, small-world analysis~\cite{watts1998}, and modularity-based community detection~\cite{newman2004,blondel2008} have been applied to biological, social, and technological networks, but not, to our knowledge, to a large formal mathematics library at the multi-layer depth we pursue here. (Blanchette et al.~\cite{blanchette2015} apply mining techniques to the Isabelle AFP, and Huch and Wenzel~\cite{huch2024} analyze the AFP's theory DAG for parallel build scheduling, but neither pursues the three-layer network analysis we develop.) In a related setting, Barbosa et al.~\cite{barbosa2013} studied the network structure of three online mathematical libraries (Wikipedia restricted to mathematics, MathWorld, and DLMF), finding large strongly connected components, the absence of clear power laws, and the effectiveness of stress centrality for navigating mathematical content. Their complex networks, however, are hyperlinked encyclopedias with informal content, while here we consider a machine-verified library in which every dependency is logically certified.
