¥subsection{Additional Empirical Observations}
¥label{sec:additional-empirical}

The graph definitions of ¥S¥ref{sec:graph-definitions} assign rich metadata to each declaration node. This section reports two completed empirical analyses that characterise the ¥emph{process} layer of the dependency graph: definitional height (a kernel-level measure of implicit coupling) and tactic usage (a proof-strategy fingerprint).

¥subsubsection{Definitional Height Distribution}
¥label{sec:definitional-height}

Lean's type checker relies on ¥emph{definitional equality}: to verify that two types match, the kernel may need to unfold a chain of definitions. For each definition, Lean computes a ¥emph{definitional height} $¥delta(d)$ (the number of regular (semireducible) definitions in its transitive dependency chain), stored in the ¥texttt{ReducibilityHints} field. This quantity is invisible in the explicit dependency graph $G_{¥mathrm{thm}}$ (it does not appear in $¥mathrm{refs}(d.¥pi)$), yet it constitutes a form of ¥emph{implicit coupling}: declarations with high‾$¥delta$ are brittle, because changes to any definition along the unfolding chain can break type-checking even if the explicit premise set is unchanged.

We extracted $¥delta(d)$ for all $101{,}021$ Mathlib definitions via the ¥texttt{DefinitionVal.hints} API. Of these, $53{,}962$ are ¥emph{regular} (semireducible) with a numeric height, $42{,}786$ are ¥emph{abbreviations} (always unfolded, no height), and $4{,}273$ are ¥emph{opaque} (never unfolded). Among regular definitions, the height distribution has median‾$7$, mean‾$10.1$, IQR‾$[4, 14]$, and a maximum of‾$60$. The distribution is roughly log-normal: $58¥%$ of definitions have height $< 10$, while a tail of $13.5¥%$ has height $¥ge 20$. This confirms that most definitions sit at shallow depth, but a non-trivial fraction involves long unfolding chains that may contribute to elaboration and compilation bottlenecks.

¥subsubsection{Tactic Usage Patterns}
¥label{sec:tactic-usage}

To extract per-declaration tactic usage, we employ ¥texttt{jixia}‾¥cite{jixia2024}, a static analysis tool for Lean‾4 developed at BICMR, Peking University (see also‾¥cite{gao2025herald}). Running ¥texttt{jixia}'s declaration plugin over all $7{,}563$ Mathlib source files yields the pretty-printed proof term for each declaration, from which we parse tactic names.

¥smallskip¥noindent¥textbf{Proof mode distribution.}
Of $235{,}586$ declarations extracted, $170{,}985$ are theorems ($72.6¥%$), $35{,}635$ definitions ($15.1¥%$), $24{,}947$ instances ($10.6¥%$), and $4{,}019$ other (abbreviations, examples, opaques, inductives).
Among theorems, $76{,}708$ ($44.9¥%$) use tactic mode (proof begins with ¥texttt{by}) and $94{,}277$ ($55.1¥%$) use term mode. An additional $3{,}202$ non-theorem declarations (primarily instances) also employ tactic proofs, bringing the total tactic-mode count to $79{,}910$.

¥smallskip¥noindent¥textbf{Proof complexity.}
Tactic-mode proofs contain $325{,}454$ tactic steps in total. The distribution of steps per proof has median‾$2$, mean‾$4.1$, and IQR‾$[1, 4]$, with a maximum of‾$173$. Single-step proofs (¥texttt{by simp}, ¥texttt{by exact ¥ldots}, etc.) account for $39.2¥%$ of all tactic proofs, and $68.7¥%$ use at most three steps. Only $9.4¥%$ require ten or more steps. This confirms that most Mathlib proofs are short: the library favors fine-grained decomposition into small lemmas over long monolithic proof scripts.

¥smallskip¥noindent¥textbf{Global tactic frequencies.}
Table‾¥ref{tab:tactic-freq} lists the fifteen most frequently invoked tactics. The top three (¥texttt{rw} ($16.9¥%$), ¥texttt{simp} ($11.2¥%$), and ¥texttt{exact} ($10.4¥%$)) together account for $38.5¥%$ of all tactic steps. The top ten account for $67.0¥%$. The dominance of ¥texttt{rw} (rewriting) reflects Mathlib's equational style: the most common proof strategy is to transform the goal by substituting known equalities. The prominence of ¥texttt{simp} (simplification) and ¥texttt{exact} (term-mode embedding) indicates that automation and direct proof-term construction are the next most common strategies.

¥begin{table}[ht]
¥centering
¥small
¥caption{Top 15 tactics by invocation count across all $79{,}910$ tactic-mode proofs ($325{,}454$ total steps).}
¥label{tab:tactic-freq}
¥begin{tabular}{l r r}
¥toprule
¥textbf{Tactic} & ¥textbf{Count} & ¥textbf{¥%} ¥¥
¥midrule
¥texttt{rw}       & 55,074 & 16.9 ¥¥
¥texttt{simp}     & 36,380 & 11.2 ¥¥
¥texttt{exact}    & 33,767 & 10.4 ¥¥
¥texttt{have}     & 23,939 &  7.4 ¥¥
¥texttt{refine}   & 18,158 &  5.6 ¥¥
¥texttt{apply}    & 12,639 &  3.9 ¥¥
¥texttt{obtain}   & 12,020 &  3.7 ¥¥
¥texttt{intro}    & 10,321 &  3.2 ¥¥
¥texttt{simpa}    &  8,372 &  2.6 ¥¥
¥texttt{ext}      &  7,408 &  2.3 ¥¥
¥texttt{rcases}   &  6,519 &  2.0 ¥¥
¥texttt{simp¥_rw} &  5,933 &  1.8 ¥¥
¥texttt{let}      &  5,439 &  1.7 ¥¥
¥texttt{rintro}   &  5,304 &  1.6 ¥¥
¥texttt{by¥_cases} & 3,959 &  1.2 ¥¥
¥midrule
¥emph{Top 3}      & 125,221 & 38.5 ¥¥
¥emph{Top 10}     & 218,070 & 67.0 ¥¥
¥bottomrule
¥end{tabular}
¥end{table}

¥smallskip¥noindent¥textbf{Per-namespace variation.}
Tactic usage varies significantly across mathematical domains. Table‾¥ref{tab:tactic-ns} shows the top tactic for each of the ten largest namespaces. Notably, ¥texttt{rw} is the most common tactic in eight of ten namespaces, but ¥texttt{simp} leads in ¥module{CategoryTheory} (where diagrammatic reasoning is heavily automated) and ¥texttt{exact} ties with ¥texttt{rw} in ¥module{Topology} (where point-set arguments are often resolved by direct term construction).

To quantify these differences, we compute the Jensen--Shannon divergence (JSD) between the tactic frequency distributions of namespace pairs. The most divergent pair is ¥module{CategoryTheory} vs.¥ ¥module{MeasureTheory} ($¥mathrm{JSD} = 0.180$), reflecting fundamentally different proof styles: categorical proofs rely heavily on ¥texttt{simp}, ¥texttt{aesop}, and ¥texttt{ext}, while measure-theoretic proofs favor ¥texttt{have}, ¥texttt{exact}, and ¥texttt{filter¥_upwards}. The most similar pairs are ¥module{Algebra} vs.¥ ¥module{LinearAlgebra} ($¥mathrm{JSD} = 0.052$) and ¥module{RingTheory} vs.¥ ¥module{LinearAlgebra} ($¥mathrm{JSD} = 0.052$), confirming that algebraic subfields share closely related proof methodologies.

¥begin{table}[ht]
¥centering
¥small
¥caption{Tactic usage by top-level namespace (top 10 by tactic-proof count). The three most frequent tactics in each namespace are listed with their share of that namespace's total tactic steps.}
¥label{tab:tactic-ns}
¥begin{tabular}{l r l}
¥toprule
¥textbf{Namespace} & ¥textbf{Proofs} & ¥textbf{Top 3 tactics (¥%)} ¥¥
¥midrule
¥module{Algebra}         & 11,390 & ¥texttt{rw} (20), ¥texttt{simp} (13), ¥texttt{exact} (9) ¥¥
¥module{Analysis}        &  9,677 & ¥texttt{rw} (16), ¥texttt{simp} (11), ¥texttt{exact} (10) ¥¥
¥module{Data}            &  8,598 & ¥texttt{rw} (18), ¥texttt{simp} (14), ¥texttt{exact} (9) ¥¥
¥module{RingTheory}      &  6,633 & ¥texttt{rw} (18), ¥texttt{simp} (10), ¥texttt{exact} (10) ¥¥
¥module{Topology}        &  5,741 & ¥texttt{rw} (14), ¥texttt{exact} (14), ¥texttt{simp} (9) ¥¥
¥module{CategoryTheory}  &  5,625 & ¥texttt{simp} (13), ¥texttt{rw} (13), ¥texttt{exact} (9) ¥¥
¥module{MeasureTheory}   &  4,631 & ¥texttt{rw} (15), ¥texttt{exact} (13), ¥texttt{have} (10) ¥¥
¥module{LinearAlgebra}   &  4,323 & ¥texttt{rw} (18), ¥texttt{simp} (12), ¥texttt{exact} (8) ¥¥
¥module{Order}           &  3,780 & ¥texttt{rw} (16), ¥texttt{simp} (14), ¥texttt{exact} (14) ¥¥
¥module{NumberTheory}    &  3,049 & ¥texttt{rw} (18), ¥texttt{have} (10), ¥texttt{exact} (10) ¥¥
¥bottomrule
¥end{tabular}
¥end{table}
