\documentclass[11pt]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{hyperref}
\usepackage{geometry}
\geometry{margin=1in}

\title{$\mathtt{Mathlib}$ as a network}
\author{Xinze Li \and Simone Severini}
\date{\today}

\begin{document}

\maketitle

\section{Introduction}

$\mathtt{Mathlib}$ is the principal mathematical library for the Lean  proof assistant~\cite{mathlib2020,lean4}. $\mathtt{Mathlib}$ comprises a growing corpus of formalized mathematics and is developed collaboratively by a global community of contributors~\cite{mathlib_community}. 

\section{Dataset}


We consider a labeled digraph.. mathlib nodes justify why we chose this graph and where the data is avaialble. 


TO DO: 

define precisely all the vertices and all the edges

define the vertices. 
theorem...
definition...
abbrev...
constructor...

347,356 arcs
edges.mathlib edges cvs 
10,417,589 

\bibliographystyle{plain}
\begin{thebibliography}{9}

\bibitem{mathlib2020}
The Mathlib Community.
\newblock The {L}ean mathematical library.
\newblock In \emph{Proceedings of the 9th ACM SIGPLAN International Conference on Certified Programs and Proofs (CPP 2020)}, pages 367--381. ACM, 2020.

\bibitem{lean4}
L.~de Moura and S.~Ullrich.
\newblock The {L}ean~4 theorem prover and programming language.
\newblock In \emph{Automated Deduction -- CADE 28}, volume 12699 of \emph{Lecture Notes in Computer Science}, pages 625--635. Springer, 2021.

\bibitem{mathlib_community}
The Mathlib Community.
\newblock Mathlib: A unified library of mathematics formalized in {L}ean.
\newblock \url{https://leanprover-community.github.io/mathlib4_docs/}, 2024.

\end{thebibliography}

\end{document}