\section{Methodology}
\label{sec:methodology}

\textbf{XXXX Maybe methodology should be before the "Our contrib" section XXXX. Here should go a small conclusion with open problems and then acknowledgments XXXX}
\xinze{This section should describe our methodology. Topics to consider: (1) Why network analysis as the analytical framework? (2) Research design: why three layers of graphs? (3) Data extraction pipeline and tooling. (4) Analysis methods and metrics chosen. (5) Reproducibility and dataset availability. Need to discuss with co-authors what belongs here vs.\ what stays in the introduction and appendix.}

\xinze{Re Methodology: most of the five topics are already covered elsewhere---(1) and (2) in the Introduction and Conceptual Framework, (3) and (5) in the Introduction's data/code availability paragraph and Appendix~B, (4) defined inline in each analysis section. \textbf{Options:} (a) keep this as a short ``bridge'' section (half a page) summarizing the three-layer design and pointing to the appendix for pipeline details; (b) remove the section entirely and distribute any remaining content to the Introduction. Leaning toward (a) for readability, but open to discussion.}

\textbf{XXXX Do we need a small conclusions session? XXXX}

\xinze{Re Conclusions: we already have a full \texttt{\\section\{Conclusion\}} (\S\ref{sec:conclusion}) with Summary, Future Directions, and Limitations.}

\paragraph{Acknowledgements.}
We thank Leonardo de Moura for facilitating the collaboration that led to this work. We are grateful to Ginestra Bianconi for insightful feedback and for drawing the analogy between cross-namespace dependencies and horizontal gene transfer in phylogenetic networks. We thank Johan Commelin for connecting us with the Mathlib maintainer community, and Oliver Nash for valuable comments on an earlier draft and for clarifications regarding Mathlib's tooling and module system.
