\section{Foundations}

Mathlib organises its source code as a collection of \texttt{.lean} files arranged in a hierarchical directory structure. Each file corresponds to a \emph{module}, and its position in the directory tree is encoded by a dot-separated name such as \texttt{Mathlib.Algebra.Group}. Because every module can \texttt{import} other modules, two natural structures coexist: a tree given by the hierarchical naming, and a directed acyclic graph given by the import relation. Our goal in this section is to formalise both structures from first principles. We begin at the coarsest level---the namespace hierarchy that mirrors the file system---and then introduce the import relation that cuts across it.

\begin{definition}[Natural numbers]
Let $\emptyset$ denote the empty set. Define the natural numbers via the von Neumann construction:
\[
  0 := \emptyset, \qquad n+1 := n \cup \{n\}.
\]
\end{definition}

\subsection*{Character encoding}

We define characters as natural numbers. The \emph{alphabet} is the set
\[
  \Sigma := \{0, 1, 2, \ldots, 127\} \subset \mathbb{N}.
\]
To interpret these numbers as characters, we adopt the standard \emph{ASCII convention}: each $n \in \Sigma$ corresponds to a specific character. For example, $65$ represents \texttt{A}, $97$ represents \texttt{a}, $48$ represents \texttt{0}, etc.

The characters relevant to Mathlib module names and their ASCII codes are:

\begin{center}
\begin{tabular}{cccccccccccc}
\toprule
Char & Code & \quad & Char & Code & \quad & Char & Code & \quad & Char & Code \\
\midrule
\texttt{A} & 65 & & \texttt{N} & 78 & & \texttt{a} & 97 & & \texttt{n} & 110 \\
\texttt{B} & 66 & & \texttt{O} & 79 & & \texttt{b} & 98 & & \texttt{o} & 111 \\
\texttt{C} & 67 & & \texttt{P} & 80 & & \texttt{c} & 99 & & \texttt{p} & 112 \\
\texttt{D} & 68 & & \texttt{Q} & 81 & & \texttt{d} & 100 & & \texttt{q} & 113 \\
\texttt{E} & 69 & & \texttt{R} & 82 & & \texttt{e} & 101 & & \texttt{r} & 114 \\
\texttt{F} & 70 & & \texttt{S} & 83 & & \texttt{f} & 102 & & \texttt{s} & 115 \\
\texttt{G} & 71 & & \texttt{T} & 84 & & \texttt{g} & 103 & & \texttt{t} & 116 \\
\texttt{H} & 72 & & \texttt{U} & 85 & & \texttt{h} & 104 & & \texttt{u} & 117 \\
\texttt{I} & 73 & & \texttt{V} & 86 & & \texttt{i} & 105 & & \texttt{v} & 118 \\
\texttt{J} & 74 & & \texttt{W} & 87 & & \texttt{j} & 106 & & \texttt{w} & 119 \\
\texttt{K} & 75 & & \texttt{X} & 88 & & \texttt{k} & 107 & & \texttt{x} & 120 \\
\texttt{L} & 76 & & \texttt{Y} & 89 & & \texttt{l} & 108 & & \texttt{y} & 121 \\
\texttt{M} & 77 & & \texttt{Z} & 90 & & \texttt{m} & 109 & & \texttt{z} & 122 \\
\midrule
\texttt{0} & 48 & & \texttt{.} & 46 & & \texttt{\_} & 95 & & & \\
\bottomrule
\end{tabular}
\end{center}

\noindent
Under this convention, the string \texttt{Algebra} corresponds to the function $s \colon 7 \to \Sigma$ with
\[
  s(0) = 65,\; s(1) = 108,\; s(2) = 103,\; s(3) = 101,\; s(4) = 98,\; s(5) = 114,\; s(6) = 97.
\]

\begin{definition}[Characters and strings]
A \emph{character} is an element of $\Sigma$. A \emph{string} of length~$n$ is a function $s \colon n \to \Sigma$, where $n = \{0, 1, \ldots, n-1\}$ is the von Neumann ordinal. That is, $s$ assigns to each position $i < n$ a character $s(i) \in \Sigma$.

For each $n \in \mathbb{N}$, let $\Sigma^n := \{s : n \to \Sigma\}$ denote the set of all strings of length $n$. The set of all finite strings is
\[
  \Sigma^* := \bigcup_{n \in \mathbb{N}} \Sigma^n.
\]
In particular, $\Sigma^0 = \{\epsilon\}$ contains only the empty string (the unique function $\emptyset \to \Sigma$).
\end{definition}

\begin{definition}[Module names]
Recall that $\Sigma^*$ is the set of all finite strings (including the empty string~$\epsilon$).
Define the set of \emph{non-empty} strings:
\[
  \Sigma^+ := \Sigma^* \setminus \{\epsilon\}.
\]
A \emph{word} is an element $w \in \Sigma^+$, i.e.\ a string of length at least one
(for example, \texttt{Mathlib}, \texttt{Algebra}, \texttt{Group}).

For a string $s \in \Sigma^*$, we denote its \textbf{length} by $|s|$, the number of characters in $s$.

Let $W = \Sigma^+$ denote the set of all non-empty strings. Define the set of \textbf{Mathlib words} as
\[
  W_{\mathrm{Mathlib}} := \{w \in W \mid w \text{ appears as a component in some Mathlib module name}\}.
\]
This is a finite set containing elements such as
\[
  \texttt{Mathlib}, \texttt{Algebra}, \texttt{Group}, \texttt{Defs}, \texttt{Ring}, \texttt{Data}, \texttt{Nat}, \texttt{Topology}, \ldots
\]
For brevity, we write $\mathcal{W} := W_{\mathrm{Mathlib}}$.

For $k \geq 1$, we write
\[
  \mathcal{W}^k := \underbrace{\mathcal{W} \times \mathcal{W} \times \cdots \times \mathcal{W}}_{k \text{ times}}
\]
for the set of all $k$-tuples of Mathlib words. Define
\[
  \mathcal{W}^+ := \bigcup_{k \geq 1} \mathcal{W}^k,
\]
the set of all non-empty finite sequences of Mathlib words.

A \textbf{module name} is a non-empty finite sequence of Mathlib words. The set of all Mathlib module names is
\[
  \mathcal{M} \subset \mathcal{W}^+ = \bigcup_{k \geq 1} \mathcal{W}^k,
\]
the subset of $\mathcal{W}^+$ consisting of module names that actually appear in the library.
We adopt the dot notation and write $w_1.w_2.\cdots.w_k$ for the tuple
$(w_1, \ldots, w_k) \in \mathcal{M}$.
For instance, the tuple $(\texttt{Mathlib},\, \texttt{Algebra},\, \texttt{Group})$
is written \texttt{Mathlib.Algebra.Group}.
\end{definition}

\begin{definition}[Prefix tree]
Define a partial order on $\mathcal{M}$: $m \leq m'$ if and only if $m$ is a prefix of~$m'$.
Then $(\mathcal{M}, \leq)$ is a rooted tree.
\end{definition}

\begin{example}
Consider the following subset of $\mathcal{M}$:
\begin{align*}
  \{\; & \texttt{Mathlib},\;
         \texttt{Mathlib.Algebra},\;
         \texttt{Mathlib.Algebra.Group}, \\
       & \texttt{Mathlib.Algebra.Ring},\;
         \texttt{Mathlib.Data},\;
         \texttt{Mathlib.Data.Nat} \;\}.
\end{align*}
The prefix order gives the tree:
\[
  \texttt{Mathlib}
  \begin{cases}
    \texttt{.Algebra}
    \begin{cases}
      \texttt{.Group} \\
      \texttt{.Ring}
    \end{cases} \\[6pt]
    \texttt{.Data}
    \begin{cases}
      \texttt{.Nat}
    \end{cases}
  \end{cases}
\]
Here $\texttt{Mathlib} \leq \texttt{Mathlib.Algebra} \leq \texttt{Mathlib.Algebra.Group}$,
but $\texttt{Mathlib.Algebra.Group}$ and $\texttt{Mathlib.Data.Nat}$ are incomparable.
\end{example}

\begin{definition}[File contents and import extraction]
Let $\mathrm{Content} \colon \mathcal{M} \to \Sigma^*$ assign to each module its source code. Define:
\[
  \mathrm{imports}(m) := \bigl\{m' \in \mathcal{M} \mid \texttt{"import } m'\texttt{"} \text{ is a substring of } \mathrm{Content}(m)\bigr\}.
\]
\end{definition}

\begin{example}
Suppose the file \texttt{Mathlib/Algebra/Group.lean} contains:
\begin{minted}[frame=single,fontsize=\small,breaklines,bgcolor=gray!5]{lean4}
import Mathlib.Algebra.Ring
import Mathlib.Data.Nat

class Group (G : Type) extends Monoid G where
  inv : G -> G
  mul_left_inv : forall a : G, inv a * a = 1
\end{minted}
Then $\mathrm{Content}(\texttt{Mathlib.Algebra.Group})$ is the entire string above.
Since it contains the substrings \texttt{"import Mathlib.Algebra.Ring"}
and \texttt{"import Mathlib.Data.Nat"}, we obtain
\[
  \mathrm{imports}(\texttt{Mathlib.Algebra.Group})
  = \{\texttt{Mathlib.Algebra.Ring},\; \texttt{Mathlib.Data.Nat}\}.
\]
\end{example}

\begin{definition}[Dependency graph]
The \emph{module dependency graph} of Mathlib is:
\[
  G = (\mathcal{M},\, E), \qquad E = \{(m, m') : m' \in \mathrm{imports}(m)\}.
\]
\end{definition}

\begin{example}
Continuing the previous example, suppose the import relations are:
\begin{itemize}
  \item \texttt{Mathlib.Algebra.Group} imports \texttt{Mathlib.Algebra.Ring} and \texttt{Mathlib.Data.Nat};
  \item \texttt{Mathlib.Algebra.Ring} imports \texttt{Mathlib.Data.Nat};
  \item \texttt{Mathlib.Data.Nat} imports \texttt{Mathlib.Data}.
\end{itemize}
Then the corresponding subgraph of~$G$ is:
\begin{center}
\begin{tikzpicture}[
    >=Stealth,
    node distance=1.8cm and 2.5cm,
    every node/.style={font=\small\ttfamily}
  ]
  \node (AG) {Algebra.Group};
  \node (AR) [below left=of AG] {Algebra.Ring};
  \node (DN) [below right=of AG] {Data.Nat};
  \node (D)  [below=of DN] {Data};
  \draw[->] (AG) -- (AR);
  \draw[->] (AG) -- (DN);
  \draw[->] (AR) -- (DN);
  \draw[->] (DN) -- (D);
\end{tikzpicture}
\end{center}
Each arrow $m \to m'$ indicates that $m$ imports~$m'$.
Note that the import edges cut across the prefix tree: \texttt{Algebra.Group}
depends on \texttt{Data.Nat}, which lies in a different branch.
\end{example}
