\section{Previous works}

\paragraph{Software dependency networks.}
Dependency graphs arise naturally in software ecosystems: package managers such as npm, PyPI, and Maven define directed acyclic graphs in which each package declares its requirements~\cite{decan2019,labelle2004}. Studies of these ecosystems have revealed heavy-tailed degree distributions, small-world connectivity, and fragility under targeted removal of hub packages. Our analysis applies a similar methodology to a formal mathematics library---a setting in which every edge carries a stronger guarantee than ``package~$A$ requires package~$B$ at build time.'' In a type-checked library, every dependency has been verified by the kernel to be logically necessary; the graph records not merely a build requirement but a semantic relationship.

\paragraph{\module{Mathlib} growth and maintenance.}
Van Doorn, Ebner, and Lewis~\cite{vandoorn2020} described the engineering practices that sustain \module{Mathlib} as a collaboratively maintained library, addressing linting, documentation, and refactoring workflows. Commelin and Topaz~\cite{commelin2023} proposed abstraction boundaries and specification-driven development as organizing principles for formal mathematics. More recently, Commelin et al.~\cite{baanen2025} documented the challenges of scaling \module{Mathlib}, including build times, namespace reorganization, and technical debt. On the tooling side, Zhu et al.~\cite{zhu2025leanarchitect} introduced \textsc{LeanArchitect}, a Lean package that extracts blueprint dependency graphs directly from Lean code, synchronizing informal \LaTeX{} exposition with formal declarations. Their work addresses the maintenance overhead of keeping blueprints consistent with evolving formalizations---a process-level concern complementary to our product-level structural analysis. These works address \module{Mathlib}'s maintenance from a software engineering perspective. Our analysis complements theirs by providing a network-theoretic diagnostic of the same challenges they address qualitatively: we show, for example, that the $17.5\%$ import redundancy rate and $0.107$ module cohesion quantify the cognitive overhead that maintenance practices must manage.

\paragraph{Machine learning for formal mathematics.}
The graph structure of \module{Mathlib} has been exploited in machine learning, particularly for premise selection. Yang et al.~\cite{leandojo} introduced the LeanDojo benchmark, extracting proof traces from \module{Mathlib} that implicitly define a dependency graph. Bemberek et al.~\cite{bemberek2025} constructed a heterogeneous dependency graph from this benchmark and applied relational graph convolutional networks for premise selection, achieving a $25\%$ improvement over text-only baselines. Lu et al.~\cite{leanfinder} built a directed acyclic graph of Lean~4 statements with dependency edges to support semantic search. On the knowledge-representation side, the EpiK Protocol~\cite{epik2024} constructed a knowledge graph from Lean~4 and \module{Mathlib}, applying embedding models for link prediction. Uskuplu and Moss~\cite{knowtex2025} argued that dependency graphs should become a standard feature of mathematical writing. These systems consume the dependency graph as training data or as a retrieval index. We study the graph itself---not as input to a downstream task, but as an object of independent structural interest.

\paragraph{Complex network methodology.}
The analytical tools we employ---degree distributions, centrality measures, community detection---originate in network science. Scale-free network theory~\cite{barabasi1999}, small-world analysis~\cite{watts1998}, and modularity-based community detection~\cite{newman2004,blondel2008} have been applied to biological, social, and technological networks, but not, to our knowledge, systematically to a large-scale formal mathematics library. In a related setting, Barbosa et al.~\cite{barbosa2013} studied the network structure of three online mathematical libraries---Wikipedia (restricted to mathematics), MathWorld, and DLMF---finding large strongly connected components, the absence of clear power laws, and the effectiveness of stress centrality for navigating mathematical content. Their objects of study, however, are hyperlinked encyclopedias with informal content; ours is a machine-verified library in which every dependency is logically certified.

\paragraph{Philosophy of mathematical practice.}
Our central thesis---that the deterministic product of formalization preserves measurable traces of the uncertain process that created it---echoes themes in the philosophy of mathematical practice. Lakatos~\cite{lakatos1976} argued that mathematical knowledge grows through a dialectical process of conjecture, proof, and refutation; Corfield~\cite{corfield2003} called for a philosophy that engages with the actual practice of working mathematicians. We share their emphasis on process over finished product, but ground it in quantitative network data rather than historical case studies. \textcolor{red}{do not include this. }

\paragraph{Gap.}
Prior work uses the dependency graph instrumentally---as training data for machine learning, as a compilation dependency for build systems, or as a navigational aid for human developers. No study has analyzed the graph itself as a structural object encoding both mathematical logic and human organizational practice. This paper takes a first step toward filling that gap.
