\section{Conclusion, Limitation, and Future Work}

We present \thiswork, a language model with 7 billion parameters that outperforms all open-source models in formal theorem proving in Lean 4.
\thiswork~is initialized with \thiswork-Base, which extends the pre-training of DeepSeekMath-Base 7B using a specialized corpus for formal mathematical reasoning.
Supervised fine-tuning is conducted on a comprehensive Lean 4 code completion dataset, encompassing a wide range of formal theorems from various mathematical domains.
Subsequently, we employ GRPO to enhance whole-proof generation through online reinforcement learning.
Upon developing the \thiswork~model, we introduce \treesearch, a variant of Monte-Carlo tree search, to improve problem-solving capabilities via large-scale search with extensive exploration.
These components form a comprehensive pipeline for training an LLM-based proof assistant, enabling \thiswork~to achieve significant improvements over \lastwork.

The framework of \thiswork~is designed to establish an AlphaZero-like pipeline for formal theorem proving.
The use of expert iteration and synthetic data mirrors the core trial-and-error loop of reinforcement learning, with the compiler oracle serving as the world model to provide environmental supervision.
Within the RL paradigm, the integrated tree search module has proven to be highly effective in advancing superhuman performance across various domains \citep{silver2016mastering, fawzi2022discovering, lutz2023top}.
In developing \thiswork, we focus on the exploration aspect of RL, introducing \treesearch~to diversify the generation of proof steps.
However, the exploitation aspect, specifically the problem of proof search, remains unexplored.
A promising future direction is training a critic model to assess incomplete proofs and prune search branches.
Such a partial-proof critic model would implicitly perform temporal credit assignment \citep{sutton1984temporal}, decomposing proof-level feedback into step-wise value differences \citep{arjona2019rudder}.
Developing critic models for assessing long planning paths and providing guidance rewards presents a crucial and challenging problem \citep{ng2000algorithms, sorg2010reward} that warrants further investigation.

Finally, recent work has progressed beyond proving individual theorems to addressing real-world theory proving within complex, multi-theorem Lean files~\citep{hu2024minictx}. This shift is a natural extension of our whole-proof generation approach. Our observations indicate that the current model already possesses some understanding of file-level context. Moving forward, we will focus on enhancing this aspect to support cutting-edge Lean mathematical formalization developers with our language model advancements.