\documentclass[11pt]{article}
\usepackage{preamble}

\title{Formalization Encodes Human Organization: 

A Network Analysis of \module{Mathlib}}
\author{Xinze Li \and Simone Severini}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
%We study the dependency structure of \module{Mathlib}, the largest unified library of formalized mathematics, comprising $7{,}563$ modules and $308{,}129$ declarations connected by $8.4$ million dependency edges. We construct dependency graphs at three granularities---modules (source files), declarations (theorems, definitions, and axioms), and namespaces (hierarchical name prefixes)---and examine how these levels relate to one another.

%The two human-facing organizational systems---file layout and namespace naming---largely agree with each other (NMI $= 0.71$) but diverge from the underlying logical dependencies. Module cohesion averages just $0.107$: roughly $90\%$ of a module's dependency traffic points outward. Louvain community detection recovers recognizable mathematical disciplines, yet their alignment with human-defined namespaces is limited (NMI $= 0.34$).

%At the declaration level, the hub structure splits into two distinct layers: a language infrastructure layer---equality reasoning, numeral coercions---that reflects Lean's type system design, and a deeper mathematical layer built on categories, real numbers, and algebraic modules. $92.2\%$ of cross-file declaration dependencies are indirect, reaching their targets through transitive import chains rather than explicit import statements. A containment analysis across namespace depths shows that human cognitive classification retains organizing power at the coarsest level ($48.9\%$ at the top-level directory) but fades to around $13\%$ at finer granularities.

We analyze the network structure of \texttt{Mathlib}, the largest unified library of formalized mathematics, to quantify the divergence between human organizational processes and logical necessity. We distinguish the \emph{module import graph}, representing file organization, from the \emph{declaration dependency graph}, representing deductive requirements. At the module level ($N \approx 7.6 \times 10^3$), the graph exhibits a deep "funnel" topology (depth 153) where 17.5\% of imports are logically redundant, serving as an aid to navigation rather than a compiler necessity. At the declaration level ($N \approx 3.1 \times 10^5$), we observe a heavy-tailed citation distribution (in-degree) contrasted with bounded proof complexity (out-degree). Centrality analysis reveals a structural stratification between language infrastructure and mathematical concepts. We find that the library forms a scale-free\xinze{Replace ``scale-free'' with ``heavy-tailed / hub-dominated''---scale-free is contested and our data shows lognormal fits better for out-degree} network that is resilient to random failure but highly vulnerable to targeted removal of hubs. While the library's organization into folders effectively separates broad fields like Algebra and Analysis, this clean hierarchy dissolves when we look at individual theorems. At that fine level, the logical connections ignore file boundaries and crisscross densely between different areas, creating dependencies that the simple file structure hides. These metrics offer a quantitative snapshot of how a large formalized library is organized, with potential applications to maintenance prioritization, import mechanism design, and premise selection in AI-assisted proof search. Our findings suggest that future proof assistants and search tools should architecturally decouple the minimal dependency graph required for verification from the redundant, semantically rich graph utilized for human interaction.
\end{abstract}

\setcounter{tocdepth}{2}
\newpage
\tableofcontents
\newpage

\input{sections/introduction}
\input{sections/module-graph}
\input{sections/declaration-graph}
\input{sections/namespace-graph}
\input{sections/structural-analysis}
\input{sections/cross-level-analysis}
\input{sections/conclusion}

\newpage
\bibliographystyle{plain}
\bibliography{references}

\newpage
\appendix
\input{sections/preliminaries}
\end{document}
