\documentclass[10pt]{article}
\usepackage{preamble}

\title{The Network Structure of \texttt{Mathlib}:\\ Software Engineering vs. Mathematical Dependencies\\[6pt]\large DRAFT V2}
\author{
  Xinze Li\thanks{Department of Mathematics, University of Toronto} 
  \and 
  Nanyun Peng\thanks{Computer Science Department, University of California, Los Angeles, and Google Cloud} 
  \and 
  Simone Severini\thanks{Department of Computer Science, University College London, and Google Cloud}
}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
We extract and analyze the dependency structure of Lean~4's \texttt{Mathlib} (308,129 declarations, 8.4 million edges) as a multi-layer network at the declaration, module, and namespace levels. We see large-scale formal libraries as \emph{transitional artifacts} that simultaneously encode inherited mathematical structure and the organizational processes of the community that builds them. The purpose of the network analysis is to disentangle these two dimensions and to determine which structural patterns constitute actionable engineering targets and which reflect intrinsic mathematical organization. To this end, we introduce graph decompositions that isolate compiler-derived edges (74.2\%) from human-authored ones, proof-driven dependencies (43.9\%) from statement-driven ones, and language infrastructure from mathematical hubs. Three principal findings emerge from this decomposition. First, human-designed taxonomies (files, namespaces) exhibit strong internal coherence yet diverge from the logical dependency structure, with cross-namespace coupling reaching 50.9\% at the declaration level. Second, the module import graph contains a layer of intentional redundancy that tooling actively manages; the \texttt{shake} linter has kept imports stable at ${\sim}3.1$ per file even as the library doubled, while the median import utilization of 1.6\% quantifies the gap between file-level and declaration-level granularity. Third, formalization preserves traditional mathematical labels but flattens their semantic content; the dependency structure is governed by the process of proving (43.9\% proof-only edges, 44.8\% terminal theorems) rather than by statement content alone. Our empirical analysis uses longitudinal data across four \texttt{Mathlib} snapshots (2024--2026), confirming that these patterns persist and that existing engineering responses (the \texttt{shake} linter, typeclass caching, parallel compilation) leave measurable traces in the network.
\end{abstract}


\addtocontents{toc}{\protect\setcounter{tocdepth}{-10}}
%\newpage
\input{sections/main/introduction}
\input{sections/main/contribution}
\input{sections/main/previous-work}
\input{sections/main/conceptual-framework}
\input{sections/main/methodology}
\input{sections/main/conclusion}

\paragraph{Acknowledgements.}
We thank Leonardo de Moura for facilitating the collaboration that led to this work, and for valuable feedback and discussions. We are grateful to Ginestra Bianconi for insightful feedback and for drawing the analogy between cross-namespace dependencies and horizontal gene transfer in phylogenetic networks. We thank Johan Commelin for connecting us with the Mathlib maintainer community, Oliver Nash for valuable comments on an earlier draft and for clarifications regarding \texttt{Mathlib}'s tooling and module system, and Franz Josef Och and Jennifer Bennett for valuable discussions. Xinze Li thanks his doctoral advisor Yevgeny Liokumovich and Luis Seco for their support, and Kasra Rafi, Bennett Chow, Bin Dong, Ronghui Gu, and Zaiwen Wen for helpful discussions on AI for mathematics.

%\clearpage
\begingroup
\footnotesize
\let\oldbibliography\thebibliography
\renewcommand{\thebibliography}[1]{%
  \oldbibliography{#1}%
  \setlength{\itemsep}{0pt}%
  \setlength{\parsep}{0pt}%
  \setlength{\parskip}{0pt}%
}
\bibliographystyle{alpha}
\bibliography{references}
\endgroup

\newpage
\addtocontents{toc}{\protect\setcounter{tocdepth}{2}}
\appendix
\begin{center}
\vspace*{2cm}
{\Huge\bfseries Appendices}
\vspace{1cm}

\end{center}
\vfill
\setcounter{tocdepth}{2}
\tableofcontents
\newpage
\input{sections/appendix/preliminaries}
\section{Dependency Graphs}
\label{sec:graph-definitions}

The dependency graphs defined in this paper are not pure embodiments of mathematical logic---each is an entanglement of inherited mathematical structure and the organizational imprint of Lean's language mechanisms. Table~\ref{tab:decompositions} catalogs the complete family of decompositions, organized by the Lean mechanism they isolate.

\begin{table}[!htb]
\centering
\caption{Tool-mediated decompositions of the dependency graphs. Each row identifies a Lean language mechanism, the corresponding graph decomposition or projection, and the dimension it isolates. Rows above the mid-rule have been extracted and analyzed; rows below are deferred to future work.}
\label{tab:decompositions}
\footnotesize
\begin{tabular}{@{}lllrl@{}}
\toprule
\textbf{Lean mechanism} & \textbf{Decomposition} & \textbf{Isolates} & \textbf{Scale} & \textbf{Ref.} \\
\midrule
\multicolumn{5}{@{}l}{\textit{Edge-level decompositions of $G_{\mathrm{thm}}$}} \\[3pt]
Statement vs.\ proof & $E_S \sqcup E_P \sqcup E_{SP}$ & Assertion vs.\ proof strategy & $8.1/43.9/48.0\%$ & Def.~\ref{def:stmt-proof-dep} \\
Explicit vs.\ synthesized & $E_{\mathrm{explicit}} \sqcup E_{\mathrm{synth}}$ & Human-written vs.\ compiler-generated & $\sigma = 74.2\%$ & Def.~\ref{def:synth-partition} \\[6pt]
\multicolumn{5}{@{}l}{\textit{Subgraphs of $G_{\mathrm{thm}}$}} \\[3pt]
Typeclass synthesis & $G_{\mathrm{tc}}$ & Algebraic hierarchy infrastructure & $1{,}817$ / $29{,}952$ & Def.~\ref{def:tc-graph} \\[6pt]
\multicolumn{5}{@{}l}{\textit{Node-level enrichment of $G_{\mathrm{thm}}$}} \\[3pt]
Attributes \& metadata & $\alpha(d)$, $\kappa(d)$, $\upsilon(d)$ & Proof complexity, polymorphism & $165$ types & Def.~\ref{def:decl-attributes} \\
\texttt{to\_additive} & $\sim_{\mathrm{add}}$ mirror pairs & Metaprogrammed duplication & $18{,}659$ pairs & Def.~\ref{def:decl-attributes} \\[6pt]
\multicolumn{5}{@{}l}{\textit{Refinements of $G_{\mathrm{module}}$}} \\[3pt]
Public/private imports & $G_{\mathrm{mod}}^{\mathrm{pub}}$ & Intentional API vs.\ all imports & $45{,}289$ / $1{,}912$ & Def.~\ref{def:visibility-graph} \\
Compilation cost & $G_{\mathrm{build}}$ & Engineering bottlenecks & $47{,}201$ edges & Def.~\ref{def:build-graph} \\
Import utilization & $\mathrm{util}(m_i, m_j)$ & Module granularity & median $1.6\%$ & Def.~\ref{def:import-util} \\[6pt]
\multicolumn{5}{@{}l}{\textit{Additional subgraphs of $G_{\mathrm{thm}}$}} \\[3pt]
Coercion insertion & $G_{\mathrm{coe}}$ & Implicit type conversion paths & $256$ coercions & Def.~\ref{def:coercion-graph} \\
Structure inheritance & $G_{\mathrm{ext}}$ & \texttt{extends} hierarchy & $1{,}535$ edges & Def.~\ref{def:extends-graph} \\
\texttt{deriving} handlers & $E_{\mathrm{auto}}$ & Auto-generated instances & $15$ handlers & Def.~\ref{def:deriving-edges} \\[6pt]
\multicolumn{5}{@{}l}{\textit{Additional node-level enrichment of $G_{\mathrm{thm}}$}} \\[3pt]
Tactic usage & $\tau(d)$ & Proof strategy selection & top: \texttt{rw}, \texttt{exact}, \texttt{simp} & Def.~\ref{def:tactic-usage} \\
Definitional height & $\delta(d)$ & Kernel-level implicit coupling & median $7$, max $60$ & Def.~\ref{def:def-height} \\
\midrule
\multicolumn{5}{@{}l}{\textit{Planned decompositions (future work)}} \\[3pt]
File co-modification & $G_{\mathrm{co}}$ & Social/development coupling & --- & \S\ref{sec:co-edit-graph} \\
Temporal evolution & $G^{(t)}$ & Structural stability over time & --- & \S\ref{sec:temporal-graphs} \\
\bottomrule
\end{tabular}
\end{table}

\noindent Each row acts as a \emph{projection} that isolates one dimension---inherited mathematical structure or human organizational process---from the entangled raw graph. The ``Scale'' column reports the footprint of each mechanism at our snapshot, extracted programmatically from Lean's \texttt{Environment} API across all $499{,}732$~Mathlib constants. All decompositions above the mid-rule have been extracted and empirically analyzed in this paper; the two remaining rows (file co-modification and temporal evolution) require additional data sources and are deferred to future work (\S\ref{sec:future-graphs}).

\input{sections/appendix/module-graph}
\input{sections/appendix/declaration-graph}
\input{sections/appendix/namespace-graph}
\input{sections/appendix/future-graphs}
\section{Network Analysis Details}\label{app:supplementary}

This section presents the detailed network statistics underlying the findings of \S\ref{sec:contribution}. The three dependency graphs defined in Appendix~\ref{sec:graph-definitions}---the module graph $G_{\mathrm{module}}$ (\S\ref{app:module-detail}), the declaration graph $G_{\mathrm{thm}}$ (\S\ref{app:decl-detail}), and the namespace graph $G_{\mathrm{ns}}$ (\S\ref{app:ns-detail})---are each examined through degree distribution, DAG depth, centrality rankings, community structure, and robustness analysis, supplemented by graph-specific measurements such as containment decay, cross-namespace heatmaps, and theorem/lemma validation. Section~\ref{sec:cross_level} then overlays the three levels, quantifying how file boundaries, namespace boundaries, and logical dependencies align and diverge.

\input{sections/appendix/appendix-module}
\input{sections/appendix/appendix-decl}
\input{sections/appendix/appendix-ns}
\input{sections/appendix/cross-level-analysis}
\end{document}
